---
title: "European Rivers' Runoff Trends"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = T)
knitr::opts_knit$set(root.dir = 'C:/Users/godzi/Documents/edafinal')
```


```{r, include = F}
# packages

library(data.table)
library(ggplot2)
library(mapview)
library(sf)
library(plyr)
library(reshape2)
library(dplyr)
getwd()
```


``` {r, include = F}
# upload files

runoff_day <- readRDS('./data/runoff_day.rds')
runoff_month <- readRDS('./data/runoff_month.rds')
runoff_stations_day <- readRDS('./data/runoff_stations_day.rds')
runoff_stations_year <- readRDS('./data/runoff_stations_year.rds')
runoff_summary <- readRDS('./data/runoff_summary.rds')
runoff_summer <- readRDS('./data/runoff_summer.rds')
runoff_winter <- readRDS('./data/runoff_winter.rds')
runoff_year_og <- readRDS('./data/runoff_year.rds')
runoff_year_river <- readRDS('./data/runoff_year_river.rds')
runoff_info <- readRDS('./data/runoff_info.rds')
runoff_year <- readRDS('./data/runoff_stations_year.rds')
runoff_rivers <- readRDS('./data/runoff_rivers.rds')
runoff_rivers_summer <- readRDS( './data/runoff_rivers_summer.rds')
runoff_rivers_winter <- readRDS( './data/runoff_rivers_winter.rds')
runoff_day_countries <- readRDS('./data/runoff_day_countries.rds')


```


```{r,include=F}

color_set <- c('#A85CCE', '#5E5CCE', '#6CA6DE', '#6DD9B2', '#6DD972', '#C3E558')

```

We have been provided with runoff data across European countries spanning since the 1800s. Due to data variability between different rivers, and to focus on only the most relevant data, we will only look at data from 1980 onwards. Furthermore, we will choose only five rivers and countries to focus on, to avoid overly large and confusing charts. Let's have a look at a map to have a better understanding of the spread of our data: 

```{r, echo = F}
# map of stations

stations_coords <- st_as_sf(runoff_info,
                                  coords = c('Lon', 'Lat'),
                            crs = 4326)

themap <- mapview(stations_coords, map.types = 'Stamen.TerrainBackground')
themap

```

From this map we can see that the highest concentration of stations are in Nordic countries and Germany. The data is therefore mostly focused on central and Northern Europe, with distinct exceptions in the Baltic States, the Balkans, Russia and the UK. 

To get a deeper understanding of the division, we'll break down the distibution of data across countries, rivers and stations.

```{r, echo = F, eval = T}

stations <- data.frame(
  category=c('CH', 'CZ', 'DE', 'DK', 'FI', 'FR', 'GB', 'HU', 'IS', 'LT', 'NL', 'NO', 'PL', 'RO', 'RS', 'RU', 'SE', 'SK', 'UA'),
  count = c(22,5,78,5,16,2,5,1,1,1,3,26,1,3,2,7,27,2,1)
)

stations$fraction <- stations$count / sum(stations$count)

stations$ymax <- cumsum(stations$fraction)
stations$ymin <- c(0, head(stations$ymax, n = -1))

theplot <- ggplot(stations, aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = category)) +
  geom_rect() +
  coord_polar(theta = 'y') +
  xlim(c(2,4))

theplot

```


Based on this distribution, we will choose to focus our exploration on Switzerland, Germany, Finland, Norway and Sweden, as they have the most stations and therefore the most data. The below plots will show us some overarching trends of runoff: 


```{r, echo = F, eval = T}

# five countries, annual low, medium, high

runoff_year_country <- runoff_year[Country == 'CH' | Country == 'DE' | Country == 'FI' | Country == 'SE' | Country == 'NO']

#LQ
ggplot(runoff_year_country[Year > 1980 & Year < 2011], aes(factor(Year), LQ)) +
  geom_boxplot(fill = color_set[4]) +
  facet_wrap(~Country, scales = 'free') +
  theme_bw() +
  ggtitle('Runoff per Country - Low Flow')

#MQ
ggplot(runoff_year_country[Year > 1980 & Year < 2011], aes(factor(Year), MQ)) +
  geom_boxplot(fill = color_set[1]) +
  facet_wrap(~Country, scales = 'free') +
  theme_bw() +
  ggtitle('Runoff per Country - Mean Flow')

#HQ
ggplot(runoff_year_country[Year > 1980 & Year < 2011], aes(factor(Year), HQ)) +
  geom_boxplot(fill = color_set[6]) +
  facet_wrap(~Country, scales = 'free') +
  theme_bw() +
  ggtitle('Runoff per Country - High Flow')



```

Looking at these graphs, we can see how in the period between 1980 and 2010 yearly river flow has varied a lot. However, better than looking at yearly data alone, we will next look at seasonal river runoff. This should give us a clearer picture of changes in the water flow over time which might otherwise average out on a yearly basis. Seasonal changes in runoff are a significant risk to water availability and security, so checking on summer and winter runoff separately should give us a better idea. 

Furthermore, we will shift our focus from countries to the five biggest rivers. This is because river data will be more significant for us to get a better understanding of the situation across the contitent, since rivers span across numerous countries. For this purpore we chose the five major rivers in the data set. These are Danube, Oder, Elbe, Rhine and Rhone. Some rivers, for example Loire, had missing data due to which we had to discard them. 


```{r, echo = F, eval=T}
# summer and winter trends, country

ggplot(runoff_rivers_summer[year > 1980], aes(x = year, y = value)) +
  geom_line(col = color_set[3]) +
  geom_point(col = color_set[5]) +
  facet_wrap(~River, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = color_set[1]) +
  geom_smooth(method= 'loess', formula = y~x, se = 0, col = color_set[2]) +
  scale_color_manual(values = color_set[c(4, 5, 6)]) +
  xlab('Year') +
  ylab('Runoff (m3/s)')+
  theme_bw() +
  ggtitle('Summer Runoff')

ggplot(runoff_rivers_winter[year > 1980], aes(x = year, y = value)) +
  geom_line(col = color_set[2]) +
  geom_point(col = color_set[3]) +
  facet_wrap(~River, scales = 'free') +
  geom_smooth(method = 'lm', formula = y~x, se = 0, col = color_set[1]) +
  geom_smooth(method= 'loess', formula = y~x, se = 0, col = color_set[6]) +
  scale_color_manual(values = color_set[c(4, 5, 6)]) +
  xlab('Year') +
  ylab('Runoff (m3/s)')+
  theme_bw()+
  ggtitle('Winter Runoff')

```

The charts above show us that the river flow of these rivers has been decreasing over time. While the steepness of the change varies between rivers, the overall trend indicates a decrease water availability across the board.